\Large Generative Dynamics of Supreme Court Citations: \\ Analysis with a New Statistical Model

### 2020-05-01 Tex/paper.tex
revising covariate terms
revising dependence terms

The first covariate we incorporate into the model accounts for the degree to which cases cite those that are similar in terms of the ideological positions of the justices who supported the decision. We account for this effect following \citet{spriggs2001explaining}, who find that cases are more likely to be overruled when the Court is ideologically distant from the median justice in the majority coalition that decided the case. \citet{clark2010locating} estimates a latent coordinate model of Supreme Court opinions based on the network of case-to-case citations. They find that the majority opinion falls at the ideal point of the median member of the majority coalition in the case. We include a covariate term in which $X_{ij}$ is the absolute difference between the Martin-Quinn scores \citep{martin2002dynamic} of the median justices in the majority coalitions for cases $i$ and $j$. We expect this variable to have a negative effect, which would correspond to cases citing those to which they are ideologically similar. In addition to the absolute difference variable, we also include a linear effect of the Martin-Quinn score of the median justice in the majority coalition for case $j$, which adjusts for constant ideological biases exhibited within a given term. %WE COULD USE THIS DATASET AS A SECOND APPLICATION

@@

The common dependence effects for these four dynamics are the number of triangles (case $i$ cites cases $j$ and $k$, and case $j$ cites case $k$), the number of mutual dyads (case $i$ cites $j$ and vice versa), the number of out-2-stars (case $i$ cites $j$ and $k$), and the number of in-2-stars (case $i$ is being cited by cases $j$ and $k$) in $C_{\leq t}$. Unfortunately, adding the triangle, out-2-star, or in-2-star statistic causes model degeneracy \citep{Handcock.2003}. Model degeneracy is a common obstacle when fitting ERGMs and occurs when the model places most of its probability mass on just a few networks, usually the empty and the full network. In this scenario, the simulated networks are too different from the observed network, making the underlying distribution defined by the model extremely unrealistic.

% reciprocity
The statistic included in the model for reciprocity counts the number of mutual dyads (i.e., dyads in which cases $i$ cites case $j$, and case $j$ cites case $i$) in $C_{\leq t}$. In practice, this is the number of mutual dyads in $C_{\leq t}$, since mutual dyads must emerge among opinions that were written in the same term. The log odds interpretation of the reciprocity statistic is that if opinions in case $i$ cite case $j$, the log odds that case $j$ also cites $i$ change by the theta relative to the configuration in which case $j$ does not cite case $i$.

% popularity
Accounting for the popularity effect through the in-2-star statistic is prone to cause model degeneracy. In an attempt to stabilize the model against model degeneracy as well as to still capture the popularity effect, \citet{SnijdersTomA.B..2006} introduces the \textit{alternating-in-k-star} statistic, which was shown to be equivalent to the \textit{geometrically weighted indegree distribution} (gwidegree) statistic introduced by \citet{hunter2006inference}. Define $D_r(C_t, C_{<t})$ as the number of nodes with indegree $r, r \in \{0, \dots , N_t -1 \}$, where $N_t$ is the number of cases in the network at time $t$, then gwidegree represents a network's indegree distribution in a single statistic by geometrically weighting the degree distribution
\begin{equation}
h_{gwidegree}(C_{t}, C_{<t}, \lambda)= \lambda \sum_{r=1}^{N_t-1}\biggl(1-\bigl(1-\frac{1}{\lambda}\bigr)^r\biggr)D_r(C_t, C_{<t}). 
\label{gwidegree}
\end{equation}
The decay parameter $\lambda \in [0, \infty ]$ controls the decline of the weight put on $D_r(C_t, C_{<t})$ as $r$ increases which means that the higher $\lambda$ the more the statistic responds to changes for cases with a high indegree. We fixed $\lambda=1$. Adding one single edge to the network changes the indegree distribution of the network so that a case with an indegree of $s$ turns into a case with an indegree of $s + 1$. With our notation this means that $D_r$ and $D_{r+1}$ get replaced by
$D_r -1$ and $D_{r+1} +1$, while no changes for $D_\ell,
 \ell \in \{0, \dots , N_t -1 \} \setminus \{r, r-1\} $ are
made. This changes the log odds in the following way
$$\text{logit} P(C_{ij,t}=1 | C_{-ij,t}, C_{<t}) = \exp \bigl(~\theta \cdot( 1- \frac{1}{\lambda})^r \bigr).$$
As stated by \citet{Levy2016} this means for the interpretation that a negative parameter value indicates the centralization of edges while a positive parameter indicates the dispersion of edges. Since we expect highly cited cases to be more likely to be cited again, we expect the gwidegree parameter to be negative.

@@

% Transitivity
We include two statistics to account for different types of transitivity.
The first transitivity statistic calculates the number of transitive ties, $i \to j$, where there is a directed path of length two from $i$ to $j$ through at least one case $k$, and $j$ and $k$ were written during a different term than $i$. We will refer to this statistic as the \textit{different term transitivity} statistic. As illustrated in chapter 2 we theorize Supreme Court citations to pattern a high degree of transitivity. As a result, we expect positive parameter estimates for the different term statistics.

The second transitivity statistic is the \textit{geometrically weighted edgewise shared partners} (gwesp) statistic introduced by \citet{hunter2006inference}. For directed networks the edgewise shared $r$-partners statistic ($ESP_r$) can take multiple forms, but we integrate the definition where cases $i$ and $j$ that are connected by a citation, also cite $r$ common other cases. We follow the notation of \citet{Butts.2008} and call this ESP specification \textit{outgoing shared partners} (OSP). Similar to the gwidegree-statistic the gwesp-statistic incorporates a network's edgewise shared partners distribution by replacing $D_r$ by $ESP_r$.
Note that unlike the different term transitivity statistic the gwesp statistic does not require the cited cases to be written in a previous term.

>>>>

The first covariate we incorporate into the model accounts for the degree to which cases cite those that are similar in terms of the ideological positions of the justices who supported the decision. We account for this effect following \citet{spriggs2001explaining}, who find that cases are more likely to be overruled when the Court is ideologically distant from the median justice in the majority coalition that decided the case. \citet{clark2010locating} estimates a latent coordinate model of Supreme Court opinions based on the network of case-to-case citations. They find that the majority opinion falls at the ideal point of the median member of the majority coalition in the case. We include a covariate term in which $X_{ij}$ is the absolute difference between the Martin-Quinn scores \citep{martin2002dynamic} of the median justices in the majority coalitions for cases $i$ and $j$. We expect this variable to have a negative effect, which would correspond to cases citing those to which they are ideologically similar. %WE COULD USE THIS DATASET AS A SECOND APPLICATION

@@

The common dependence effects for these four dynamics are the number of triangles (case $i$ cites cases $j$ and $k$, and case $j$ cites case $k$), the number of mutual dyads (case $i$ cites $j$ and vice versa), the number of out-2-stars (case $i$ cites $j$ and $k$), and the number of in-2-stars (case $i$ is being cited by cases $j$ and $k$) in $C_{\leq t}$. Unfortunately, adding the triangle, out-2-star, or in-2-star statistic causes model degeneracy \citep{Handcock.2003}. Model degeneracy is a common obstacle when fitting ERGMs, and occurs when the model places most of its probability mass on just a few networks, usually the empty and the full network. In this scenario, the simulated networks are too different from the observed network, making the underlying distribution defined by the model extremely unrealistic.

% reciprocity
The statistic included in the model for reciprocity counts the number of mutual dyads (i.e., dyads in which cases $i$ cites case $j$, and case $j$ cites case $i$) in $C_{\leq t}$. In practice, this is the number of mutual dyads in $C_{\leq t}$, since mutual dyads must emerge among opinions that were written in the same term. The log odds interpretation of the reciprocity statistic is that if the opinion in case $i$ cites case $j$, the log odds that case $j$ also cites $i$ change by the estimated coefficient relative to the configuration in which case $j$ does not cite case $i$. We expect the reciprocity effect to be positive.

% popularity
Accounting for the popularity effect through the in-2-star statistic is prone to cause model degeneracy. In an attempt to stabilize the model against model degeneracy as well as to still capture the popularity effect, \citet{SnijdersTomA.B..2006} introduces the \textit{alternating-in-k-star} statistic, which was shown to be equivalent to the \textit{geometrically weighted indegree distribution} (gwidegree) statistic introduced by \citet{hunter2006inference}. Define $D_r(C_t, C_{<t})$ as the number of nodes with indegree $r, r \in \{0, \dots , N_t -1 \}$, where $N_t$ is the number of cases in the network at time $t$, then gwidegree represents a network's indegree distribution in a single statistic by geometrically weighting the degree distribution
\begin{equation}
h_{gwidegree}(C_{t}, C_{<t}, \lambda)= \lambda \sum_{r=1}^{N_t-1}\biggl(1-\bigl(1-\frac{1}{\lambda}\bigr)^r\biggr)D_r(C_t, C_{<t}). 
\label{gwidegree}
\end{equation}
The decay parameter $\lambda \in [0, \infty ]$ controls the decline of the weight put on $D_r(C_t, C_{<t})$ as $r$ increases, which means that the higher $\lambda$ the more the statistic responds to changes for cases with a high indegree. We fixed $\lambda=1$. Adding one single edge to the network changes the indegree distribution of the network so that a case with an indegree of $s$ turns into a case with an indegree of $s + 1$. With our notation this means that $D_r$ and $D_{r+1}$ get replaced by
$D_r -1$ and $D_{r+1} +1$, while no changes for $D_\ell,
 \ell \in \{0, \dots , N_t -1 \} \setminus \{r, r-1\} $ are
made. This changes the log odds in the following way
$$\text{logit} P(C_{ij,t}=1 | C_{-ij,t}, C_{<t}) = \exp \bigl(~\theta \cdot( 1- \frac{1}{\lambda})^r \bigr).$$
As stated by \citet{Levy2016}, this means for the interpretation that a negative parameter value indicates the centralization of edges (i.e., popularity) while a positive parameter indicates the dispersion of edges (i.e., new ties going to less popular nodes). Since we expect highly cited cases to be more likely to be cited again, we expect the gwidegree parameter to be negative.

% Transitivity
We include two statistics to account for different types of transitivity.
The first transitivity statistic calculates the number of transitive ties, $i \to j$, where there is a directed path of length two from $i$ to $j$ through at least one case $k$, and $j$ and $k$ were written during a different term than $i$. We will refer to this statistic as the \textit{different term transitivity} statistic. As illustrated in Section 2, we theorize Supreme Court citations to pattern a high degree of transitivity. As a result, we expect positive parameter estimates for the \textit{different term transitivity} statistic.

The second transitivity statistic, intended to capture clustering that includes more than one tie formed in the same term, is the \textit{geometrically weighted edgewise shared partners} (gwesp) statistic introduced by \citet{hunter2006inference}. For directed networks the edgewise shared $r$-partners statistic ($ESP_r$) can take multiple forms, but we integrate the definition where cases $i$ and $j$ that are connected by a citation, also cite $r$ common other cases. We follow the notation of \citet{Butts.2008} and call this ESP specification \textit{outgoing shared partners} (OSP). Similar to the gwidegree-statistic the gwesp-statistic incorporates a network's edgewise shared partners distribution by replacing $D_r$ by $ESP_r$. Triangles that are incident to the outgoing shared partners configuration arise when there is a cite between two cases decided at the same time, and those two cases cite one or more of the same third cases (decided either previously or also during the same term). This would happen if citations were clustering around legal issues that are not completely captured by our same-issue-area effects.  As we do not expect that the issue area variable will account for all of the clustering around legal issues, we expect the gwesp statistic to also carry a positive coefficient value.

@@
